\label{sec:2}
% Well-motivated project with success criteria well-justified.
% Challenging and well-presented background covering Comp Sci topics beyond Part IB.
% Good requirements analysis, justified selection of suitable tools, good engineering approach.

\section{Environment}
\label{sec:2.1}

To evaluate communication in this framework, we investigate the following cooperative game.

There is an $M$x$N$ board.
The game lasts for $T$ timesteps, and at each timestep, every agent simultaneously completes one action.
The goal of the game is for the agents to cooperate and complete a recipe.

The world contains entities called agents and stations.

\textbf{Stations}

For each world generated, stations are randomly placed across the map and are immovable.
This randomisation prevents agents from overfitting a single layout.
Each station has an assigned ingredient. The station can either be of type 'pick up' or 'drop off'.
A 'pick up' station only allows the agents to take ingredients from it.
A 'drop off' station only allows agents to drop off the assigned ingredient.

\textbf{Agents}

For a similar reason, agents are randomly placed across the map at world initialisation.

Each agent is assigned an ordered pair of ingredient sets ($I_\text{in}$, $I_\text{out}$).
The agent must first collect all ingredients in $I_\text{in}$.
Then the agent's inventory is replaced with $I_\text{out}$, and the agent must remove all the ingredients from their inventory.
This defines a single recipe, but agents may complete multiple recipes in their lifetime.

Ingredients can be collected and dropped off at both other agents and stations.

One important feature of this game is that the agents have a limited field of view and can only see the nearest entity that is not their own pick-up station at any given timestep.
This converts this game from a Markov Decision Process to a Partially Observable Markov Decision Process.
The agents must work together to discover the required stations and use them to complete recipes.

At each timestep, every agent takes a single action.
Agents have five possible actions: move north, move south, move west, move east and interact.
The first four of which move the agent one step in the given direction.
If agent A executes the 'Interact' action:
\begin{enumerate}
    \item If visible agent B has a required item, place it in A's inventory, and remove it from B's inventory.
    \item If visible 'pick up' station has a required item, place in A's inventory.
    \item If visible 'drop off' station wants an item we have, remove from A's inventory.
\end{enumerate}
These five actions form the action space $\mathcal{A}$ in the POMDP definition (see section \ref{sec:2.2}).

\begin{figure}[h]
    \centering
    \begin{tikzpicture}

    \def\M{8}
    \def\N{6}

    \draw[step=1cm, lightgray, thick] (0,0) grid (\M,\N);

    \foreach \x in {1,...,\M} \node[anchor=north, text=gray] at (\x-0.5, 0) {\x};
    \foreach \y in {1,...,\N} \node[anchor=east, text=gray] at (0, \y-0.5) {\y};

    \tikzset{
        agent/.style={circle, draw=black, thick, fill=blue!60, minimum size=0.6cm, text=white, font=\sffamily\bfseries},
        pickup/.style={rectangle, draw=black, thick, fill=green!60!black, minimum size=0.8cm, text=white, font=\sffamily\bfseries},
        dropoff/.style={rectangle, draw=black, thick, fill=red!70!black, minimum size=0.8cm, text=white, font=\sffamily\bfseries},
        sight/.style={dashed, draw=blue, fill=blue!10, opacity=0.3}
    }

    \node[pickup] at (1.5, 4.5) {P1};
    \node[pickup] at (6.5, 1.5) {P2};
    \node[dropoff] at (4.5, 5.5) {D1};
    \node[dropoff] at (2.5, 1.5) {D2};

    \fill[sight] (2.5, 3.5) circle (1.45cm);
    \draw[dashed, blue] (2.5, 3.5) circle (1.45cm);
    \node[agent] at (2.5, 3.5) {A1};

    \fill[sight] (5.5, 3.5) circle (1.45cm);
    \draw[dashed, blue] (5.5, 3.5) circle (1.45cm);
    \node[agent] at (5.5, 3.5) {A2};

    \begin{scope}[shift={(8.5, 4.8)}]
        \draw[thick, rounded corners, fill=gray!5] (0,-4) rectangle (6.0, 0.5);
        \node[anchor=west, font=\sffamily\bfseries] at (0.2, 0) {Key};
        
        \node[agent, scale=0.8] at (0.6, -0.8) {};
        \node[anchor=west, font=\sffamily] at (1.0, -0.8) {Agent};
        
        \node[pickup, scale=0.8] at (0.6, -1.6) {};
        \node[anchor=west, font=\sffamily] at (1.0, -1.6) {Pick-up Station};
        
        \node[dropoff, scale=0.8] at (0.6, -2.4) {};
        \node[anchor=west, font=\sffamily] at (1.0, -2.4) {Drop-off Station};
        
        \fill[sight] (0.6, -3.2) circle (0.3cm);
        \draw[dashed, blue] (0.6, -3.2) circle (0.3cm);
        \node[anchor=west, font=\sffamily] at (1.0, -3.2) {Range of Sight};
    \end{scope}

    \end{tikzpicture}
    \caption{Example World}
    \label{fig:example_world}
\end{figure}

Figure \ref{fig:example_world} shows an example of a typical world where there are two agents and two stations.
Here, each agent's own pick-up station is visible only to the other if within the limited field of view.
If no entity is within the agent's range of view, then the agent views a 'no-entity' token.
Without communication, both agents would have to search the map independently until they discovered the station.
With communication, the agents can tell one another the location of the required pick-up station.
This shows how the environment creates asymmetric information and how communication can solve it.

\section{MARL background}
\label{sec:2.2}

Reinforcement Learning (RL) differs from classical supervised and unsupervised learning because we do not have all the data before training.
We must build environments that allow agents to take actions and receive feedback from those actions.
The agents must learn to maximise the total reward collected over time.

Classic RL is modelled as a Markov Decision Process (MDP), but to investigate communication, we must limit each agent's information, so they must communicate missing information to one another.
Therefore, we will model this RL as a Partially Observable Markov Decision Process (POMDP).

A POMDP is formally defined as $\langle \mathcal{S}, \mathcal{A}, \mathcal{P}, \mathcal{R}, \Omega, \mathcal{O}, \gamma \rangle$.
Where:
\begin{itemize}
    \item $\mathcal{S}$ is the set of all possible states of the environment.
    \item $\mathcal{A}$ is the action space. \{North, South, East, West, Interact\}.
    \item $\mathcal{P}$ is the transition probability function $P(s'|s,a)$, representing the probability that taking action $a$ in state $s$ will lead to state $s'$.
    \item $\mathcal{R}$ is the reward function. See section \ref{sec:3.3.1}.
    \item $\Omega$ is the set of all possible local observations.
    \item $\mathcal{O}$ is the observation function $P(o|s,a)$, representing the agent seeing observation $o$ given state $s$ and previous action was $a$.
    \item $\gamma$ is the discount factor. Set at 0.99 following \cite{ppo_bestpractices}.
\end{itemize}

For the environment described in section \ref{sec:2.1}, $\mathcal{S}$ consists of the world layout including the entities' locations, the agents' inventories, the station type and the station's resources.
$\Omega$ contains each agent's nearest entity if within the range of view, communication codes all other agents emitted in the previous timestep, and their own state (i.e. location and inventory).
Therefore, the system is a cooperative Dec-POMDP in which multiple agents act on the same state, receive individual local observations, and share a common reward signal.

This research's focus is emergent communication in MARL; however, MARL is often subject to various challenges that must first be overcome before the communication can be analysed.
One such problem is non-stationarity, which arises when multiple agents learn simultaneously, creating an environment that changes unpredictably from an individual agent's perspective.
Another challenge is the credit assignment problem, which arises when multiple agents work cooperatively toward a single goal, making it difficult to determine each agent's contribution.
To mitigate these problems, an appropriate, well-tested algorithm is chosen.

% The framework being implemented in the project is built on the standard centralised training, decentralised execution (CTDE) paradigm for cooperative MARL.
Multi-Agent Proximal Policy Optimisation (MAPPO) \cite{MAPPO} was selected, and alternatives were actively rejected.

Value-Decomposition Networks (VDN) \cite{VDN}, such as QMIX \cite{QMIX}, are unsuitable for the given game.
QMIX relies on a monotonicity assumption (the global optimal action-value is monotonic for each individual agent's action-values).
For training communication, the utility of a code broadcast by agents depends on the environment state, which directly violates the monotonicity assumption.

Off-policy actor-critic methods such as Multi-Agent Deep Deterministic Policy Gradient (MADDPG) \cite{MADDPG} are also unsuitable, as code broadcast semantically shifts during training, leading to new policies being trained on stale data.

MAPPO avoids these issues \cite{MAPPO}.
MAPPO is an extension of Proximal Policy Optimisation (PPO) \cite{PPO} to cooperative multi-agent settings, itself derived from Trust Region Policy Optimisation (TRPO)\cite{TRPO}.
MAPPO shares a policy across all agents and uses a centralised value function with access to the global state during training, whilst each agent acts in a decentralised manner at execution time.
This centralised training, decentralised execution paradigm is well-suited for emergent communication.
The centralised critic during training helps mitigate non-stationarity and credit assignment problems, separating them from communication problems by learning from joint states and rewards, whilst PPO-style clipped updates provide stable policy optimisation.
All agents sharing a single policy but acting independently at execution time ensure that any performance gains arise from useful information in the communication channel rather than from hidden, centralised information.

MAPPO optimises two main components: each agent's Policy (Actor) ${\pi}_{{\theta}_{i}}$ and the Value Function (Critic) $V_{\phi}$.

\subsubsection{Actor: Policy Optimisation}
\label{sec:2.2.1}
The policy network is parameterised by $\theta$ whose goal is to maximise the clipped surrogate objective, preventing drastic changes to the policy from single timesteps.
For every agent $i$, define the probability ratio $r_i(\theta)$ as:

\begin{equation}
r_i(\theta) = \frac{\pi_{\theta}(a_i | o_i)}{\pi_{\theta_{old}}(a_i | o_i)}
\end{equation}

The MAPPO objective function is:

\begin{equation}
L(\theta) = \mathbb{E}_{i} \left[ \min \left( r_i(\theta) \hat{A}_i,\ \operatorname{clip}(r_i(\theta), 1-\epsilon, 1+\epsilon) \hat{A}_i \right) \right]
\end{equation}

Where:
\begin{itemize}
\item $a_i$ is the action taken by agent $i$.
\item $o_i$ is the local observation of agent $i$.
\item $\pi_{\theta}(a_i | o_i)$ is the current policy's probability of taking action $a_i$ given observation $o_i$.
\item $\pi_{\theta_{old}}(a_i | o_i)$ is the old policy's probability of taking the same action prior to the update.
\item $\mathbb{E}_{i}$ denotes the empirical expectation over a batch of agent experiences.
\item $\hat{A}_i$ is the Advantage Estimation for agent $i$ (detailed later in Section \ref{sec:2.2}).
\item $\epsilon$ is a hyperparameter that limits the policy update.
\end{itemize}

\subsubsection{Critic: Value Function Learning}
\label{sec:2.2.2}

Unlike standard Proximal Policy Optimisation (PPO)\cite{PPO}, the Value Function is not per-agent, but per-world.
It has access to global information.
The critic network is parameterised by $\phi$ whose goal is to minimise the Mean Squared Error between its own prediction and the discounted returns $R_t$.

\begin{equation}
L(\phi) = \mathbb{E}_{t} \left[ (V_{\phi}(S_t) - R_t)^2 \right]
\end{equation}

Where:
\begin{itemize}
\item $\mathbb{E}_{t}$ denotes the expectation over timesteps $t$ within the sampled trajectories.
\item $S_t$ is the global state of the environment at timestep $t$.
\item $V_{\phi}(S_t)$ is the critic network's estimate of the value of the global state $S_t$.
\end{itemize}

\textbf{Generalised Advantage Estimation}
\label{sec:2.2.3}

MAPPO uses Generalised Advantage Estimation \cite{GAE} to balance bias and variance.
To calculate, we first find the Temporal Difference (TD) Residual (representing the error in our current estimate of the state value):

\begin{equation}
\delta_t = r_t + \gamma V(s_{t+1}) - V(s_t)
\end{equation}

We then take an exponentially weighted moving average of our estimated advantage over different time horizons.

\begin{equation}
\hat{A}_t = \sum_{l=0}^{\infty} (\gamma \lambda)^l \delta_{t+l}
\end{equation}

Where:
\begin{itemize}
\item $\delta_t$ is the TD residual at timestep $t$.
\item $r_t$ is the immediate reward received at timestep $t$.
\item $\lambda$ is the GAE trace-decay parameter ($0 \leq \lambda \leq 1$) that smooths the bias-variance trade-off.
\item $l$ is the forward-looking step index.
\end{itemize}


\section{Emergent Communication in MARL}
\label{sec:2.3}

Early work in emergent communication studied how agents can learn to communicate with one another without concern for interoperability or bandwidth, typically in two-agent coordination tasks \cite{InitialEmergent}.
This established that a communication channel can emerge purely from task reward, but they often used continuous communication.

Foundational protocols such as Differentiable Inter-Agent Learning (DIAL) \cite{dial-rial} enabled continuous gradient flow between agents during training.
This relied on passing high-dimensional, floating-point vectors, which in reality is unreasonable due to limited bandwidth.
Later work, such as Targeted Multi-Agent Communication (TarMAC) \cite{tarmac}, attempted to address this bandwidth issue by employing attention mechanisms to learn which agent to communicate with.
However, TarMAC still relied on passing high-dimensional, floating-point vectors, which remain neither realistic nor interpretable.

More recent research in multi-agent reinforcement learning studies agents learning to send discrete symbols, showing that discrete channels can match or outperform continuous channels, whilst also allowing for interoperability \cite{DiscreteComms}.
With this interoperability, the trust deficit in safety-critical applications can be reduced and even allow human-in-the-loop settings.

However, a challenge in learning discrete communication on-policy is the "Joint Exploration Problem" and the resulting "Communication Vacuum Equilibrium", as stated by Eccles et al\cite{Biases}.
At the start of training, the agent's communication policy outputs random values, providing no predictive value.
Therefore, the section of the agent's network which listens to this noise rapidly updates its weights to silence it.
However, once the weights become zero, the gradients stop flowing through the communication channel, trapping the agents in the Communication Vacuum Equilibrium.

Several works attempt to address this Communication Vacuum Equilibrium by introducing architectural biases or auxiliary objectives (e.g., adding rewards for mutual information between messages and observations).
Whilst these approaches improve communication usage, they reduce generality.
Liu et al. propose the AI Mother Tongue \cite{AIMotherTongue} architecture, in which language is learned offline, separate from online reinforcement learning.
Then the agents are trained in this language using auxiliary losses rather than relying on architectural biases or auxiliary objectives.
This project adopts this architecture and expands both the evaluation and task to determine if robust communication can emerge.
% \subsection{Communication Protocols}
% \label{sec:2.1.3}

% \subsubsection{No Communication}
% \label{sec:2.1.3.1}

% To understand the difference between optimal movement policies and the advantage of communication, a protocol is developed without communication.

% \subsubsection{On-Policy Fixed Discrete Communication}
% \label{sec:2.1.3.2}

% Another comparable baseline is required, one with communication.
% For this, each agent broadcasts a set of discrete values learned jointly with its policy at every timestep, and is then fed to all other agents at the next timestep.
% The sender's symbols are trained solely using task reward (no auxiliary losses or offline pretraining).
% This baseline captures the typical on-policy setting that suffers from the "Joint Exploration Problem", and so serves as a natural comparison for the AI Mother Tongue Architecture.

\section{AI Mother Tongue Architecture}
\label{sec:2.4}

The AI Mother Tongue \cite{AIMotherTongue} architecture addresses the "Joint Exploration Problem" by decoupling the emergence of discrete symbols from the reinforcement learning training and adding auxiliary losses as described in figure \ref{fig:ai-mother-tongue}.
Liu et al. report promising phenomena such as sudden semantic compression and interpretable symbol usage, but their evaluation is restricted to a simple cooperative task.
The underlying discrete bottleneck uses Vector-Quantised Variational Autoencoders (VQ-VAE) \cite{VQVAE}, but Liu et al. propose more expressive codebook learning methods, such as Stochastically-Quantised Variational Autoencoders (SQ-VAE) \cite{SQVAE}, and ways to advance the language, such as using Hierarchically-Quantised Variational Autoencoders (HQ-VAE) \cite{HQVAE}.


\begin{figure}[htbp]
    \centering

    \resizebox{\textwidth}{!}{
    \begin{tikzpicture}[
        >=stealth,
        node distance=1.5cm and 0.8cm,
        box/.style={draw, rectangle, rounded corners, align=center, minimum height=1.2cm, minimum width=2.4cm, fill=blue!10, font=\sffamily},
        actor/.style={draw, rectangle, rounded corners, align=center, minimum height=1.5cm, minimum width=2.8cm, fill=orange!20, font=\sffamily\bfseries},
        frozen/.style={actor, fill=gray!20, thick},
        action/.style={draw, rectangle, rounded corners, align=center, minimum height=1cm, minimum width=3.2cm, font=\sffamily},
        txt/.style={align=center, font=\sffamily},
        line/.style={draw, ->, thick}
    ]

    \node[txt] (pre_in) {Observations};
    \node[box, right=of pre_in] (pre_enc) {Encoder};
    \node[box, right=of pre_enc, fill=red!15] (pre_cb) {Quantisation};
    \node[box, right=of pre_cb] (pre_dec) {Decoder};
    \node[txt, right=of pre_dec] (pre_out) {Reconstruction};

    \draw[line] (pre_in) -- (pre_enc);
    \draw[line] (pre_enc) -- (pre_cb);
    \draw[line] (pre_cb) -- (pre_dec);
    \draw[line] (pre_dec) -- (pre_out);

    \node[above=0.8cm of pre_cb, font=\sffamily\bfseries\large] (pre_title) {Phase 1: VQ-VAE Training Loop};

    \begin{scope}[on background layer]
        \node[draw, dashed, thick, fill=blue!5, rounded corners, inner sep=15pt, fit=(pre_in)(pre_out)(pre_title)] (pre_box) {};
    \end{scope}

    \node[txt, below=4.5cm of pre_in, xshift=-0.4cm] (obs) {Observations};

    \node[frozen, right=0.5cm of obs] (frozen_enc) {Frozen Encoder\\+ Quantiser};

    \draw[line] (obs) -- (frozen_enc);
        
    \node[actor, right=1cm of frozen_enc] (actor_net) {Actor Network\\(Policy $\pi_\theta$)};
    
    \draw[line] (frozen_enc.east) -- (actor_net.west);

    \node[action, fill=purple!15, above right=0.6cm and 1cm of actor_net] (out_comm) {Codebook\\(from pretrained\\VQ-VAE)};
    \node[action, fill=teal!15, right=1cm of actor_net] (out_act) {Action};
    \node[action, fill=yellow!15, below right=0.6cm and 1cm of actor_net] (out_aux) {Auxiliary Losses};

    \draw[line] (actor_net.east) -- ++(0.5, 0) |- (out_comm.west);
    \draw[line] (actor_net.east) -- (out_act.west);
    \draw[line] (actor_net.east) -- ++(0.5, 0) |- (out_aux.west);

    \node[above=2.4cm of actor_net, xshift=-1.6cm, font=\sffamily\bfseries\large] (actor_title) {Phase 2: RL Training (AI Mother Tongue Setup)};

    \end{tikzpicture}
    }
    \caption{Architecture diagram for frozen Vector-Quantised Variational Autoencoder (VQ-VAE) pretraining and RL training the actor network.}
    \label{fig:ai-mother-tongue}
\end{figure}

Figure \ref{fig:ai-mother-tongue} shows the architecture of the AI Mother Tongue protocol proposed by Liu et al.\cite{AIMotherTongue}
Phase 1 is trained offline using collected observations, and the encoder and codebook (i.e., all possible quantised values) are saved.
Liu et al. explicitly state values indicating a healthy encoder and codebook (reconstruction loss between 0.08 and 0.15, and commitment loss below 1.0).
This dissertation will evaluate the health of the trained codebooks in section \ref{sec:4.3}.

At the initialisation of phase 2, the encoder and codebook are loaded, frozen, and used for the actor network's training.
The architecture has three outputs:
\begin{itemize}
    \item \textbf{Codebook}
    A code is emitted from the pretrained codebook chosen by a Multi-Layer Perceptron (MLP).
    \item \textbf{Action Head}
    Liu et al. use a simple linear rule as the action head based on the code emitted by the architecture and a pre-selected value $K$.
    \begin{equation}
        \text{Action}(code) =
        \begin{cases}
            \text{Action 1}, & \text{if code} < \frac{K}{2} \\
            \text{Action 2}, & \text{otherwise}
        \end{cases}
    \end{equation}
    \item \textbf{Auxiliary Head}
    These heads predict the value or likely effect of the emitted codes, helping agents learn which codes are useful and how other agents may respond.
    The auxiliary heads provide additional learning signals that are backpropagated into the actor network to guide and improve policy optimisation.
\end{itemize}

This dissertation adopts the AI Mother Tongue architecture but extends it to a more complex MAPPO setting.
Instead of a simple linear action head used by Liu et al., a Multi-Layer Perceptron (MLP) is used.
The actor network will also contain a Long Short-Term Memory (LSTM) module.
In addition to VQ-VAE, this project evaluates alternative discrete bottlenecks such as SQ-VAE and HQ-VAE (see Appendix \ref{sec:hq-vae}).

Liu et at. propose that the "Joint Exploration Problem" is solved by both the semantic bottleneck and the reflection strategies; however, evaluates the effect of neither of them individually.
This dissertation hypothesises that the complex semantic bottleneck may, in fact, be impeding the agents' learning, and that applying the reflection strategies to a standard on-policy discrete communication protocol may yield a more efficient novel solution.
% \section{MARL Algorithms}
% \label{sec:2.3}

% The framework being implemented in the project is built on the standard centralised training, decentralised execution (CTDE) paradigm for cooperative MARL.
% Multi-Agent Proximal Policy Optimisation (MAPPO) \cite{MAPPO} was selected, and alternatives were actively rejected.

% Value-Decomposition Networks (VDN) \cite{VDN}, such as QMIX \cite{QMIX}, are unsuitable for the given game.
% QMIX relies on a monotonicity assumption (the global optimal action-value is monotonic for each individual agent's action-values).
% For training communication, the utility of a code broadcast from agents has a value contextually dependent on the environment state, which directly violates the monotonicity assumption.

% Off-policy actor-critic methods such as Multi-Agent Deep Deterministic Policy Gradient (MADDPG) \cite{MADDPG} are also unsuitable as codes broadcast semantically shift during training, leading to new policies being trained on stale data.

% MAPPO avoids these issues \cite{MAPPO}.
% MAPPO evaluates the code's current semantic meaning, eliminating training on stale code.
% Furthermore, the clipped surrogate objective of the PPO algorithm is used, preventing any aggressive gradient updates, causing forgetting of the semantic meaning of codes.




% \section{Related Work}
% \label{sec:2.2}

% \subsection{Emergent Communication in MARL}
% \label{sec:2.2.1}

% Early work in emergent communication studied how agents can learn to communicate with one another without a concern for interoperability and bandwidth \cite{InitialEmergent}.
% More recent research in multi-agent reinforcement learning studies agents learning to send discrete symbols, showing that discrete channels can match or outperform continuous channels, whilst also allowing for interoperability \cite{DiscreteComms}.
% A challenge in learning discrete communication on-policy is the "Joint Exploration Problem" and the resulting "Communication Vacuum Problem", as stated by Eccles et al\cite{Biases}, where agents learn to ignore the communication channel.

% \subsection{AI Mother Tongue and VAE-based symbol systems}
% \label{sec:2.2.2}

% The AI Mother Tongue \cite{AIMotherTongue} architecture addresses the "Joint Exploration Problem" by decoupling the emergence of discrete symbols from the reinforcement learning training and adding auxiliary losses as described in figure \ref{fig:ai-mother-tongue}.
% Liu et al. report promising phenomena such as sudden semantic compression and interpretable symbol usage, but their evaluation is restricted to a simple cooperative task.
% The underlying discrete bottleneck uses Vector-Quantised Variational Autoencoders (VQ-VAE) \cite{VQVAE}, but Liu et al. propose more expressive codebook learning methods, such as Stochastically-Quantised Variational Autoencoders (SQ-VAE) \cite{SQVAE}, and ways to advance the language, such as using Hierarchically-Quantised Variational Autoencoders (HQ-VAE) \cite{HQVAE}.

% \subsection{MARL algorithms \& MAPPO}
% \label{sec:2.2.3}

% % The framework being implemented in the project is built on the standard centralised training, decentralised execution (CTDE) paradigm for cooperative MARL.
% % Multi-Agent Proximal Policy Optimisation (MAPPO) \cite{MAPPO} is chosen as empirical studies have shown surprisingly competitive results when compared with more complex off-policy and value-decomposition networks \cite{MAPPO}.
% % Alternatives such as QMIX or MADDPG offer different trade-offs but ultimately introduce extra architectural complexity, which may complicate the communication protocol evaluation.

% The framework being implemented in the project is built on the standard centralised training, decentralised execution (CTDE) paradigm for cooperative MARL.
% Multi-Agent Proximal Policy Optimisation (MAPPO) \cite{MAPPO} was selected, and alternatives were actively rejected.

% Value-Decomposition Networks (VDN) \cite{VDN}, such as QMIX \cite{QMIX}, are unsuitable for the given game.
% QMIX relies on a monotonicity assumption (the global optimal action-value is monotonic for each individual agent's action-values).
% For training communication, the utility of a code broadcast from agents has a value contextually dependent on the environment state, which directly violates the monotonicity assumption.

% Off-policy actor-critic methods such as Multi-Agent Deep Deterministic Policy Gradient (MADDPG) \cite{MADDPG} are also unsuitable as codes broadcast semantically shift during training, leading to new policies being trained on stale data.

% MAPPO avoids these issues \cite{MAPPO}.
% MAPPO evaluates the code's current semantic meaning, eliminating training on stale code.
% Furthermore, the clipped surrogate objective of the PPO algorithm is used, preventing any aggressive gradient updates, causing forgetting of the semantic meaning of codes.

\section{Requirements Analysis}
\label{sec:2.5}

\subsection[Success Criteria]{Success Criteria\footnotemark}
\label{sec:2.5.1}
\footnotetext{These success criteria restate the content of the project proposal in a more structured form.}

% \begin{enumerate}
%     \item Design and implement a vectorised MARL simulation in Rust to train and demo multi-agent interactions capable of achieving a minimum throughput of 50,000 environment steps per second (SPS) to ensure sufficient collection of training data. The simulation should be able to have single runs visualised.
%     \item Implement a functional MAPPO-based multi-agent reinforcement learning (MARL) system capable of learning, given a simulation program with set methods.
%     \item Develop a communication protocol that allows agents to transmit raw data between each other.
%     \item Implement the AI Mother Tongue architecture, demonstrating ideal reconstruction loss and commitment cost defined by Liu et al.
%     \item The training and evaluation pipeline should be able to train and evaluate models automatically and allow interchangeably with any communication protocols. Automated checkpointing and logging must also run alongside training.
% \end{enumerate}

% \subsection{Extension}
% \label{sec:2.3.2}

% \begin{itemize}
%     \item Implement a third communication protocol using a hierarchically quantised variational autoencoder (HQ-VAE). This method was suggested by the AI Mother Tongue paper as a possible future work.
%     \item Add hyperparameter tuning for the AI Mother Tongue architecture. This can involve the language size, quantisation levels, and message frequency to optimise agent coordination and learning efficiency.
%     \item Add a new task where agents must solve for objectives other than an assembly line (e.g., coordinated navigation, resource gathering, or adversarial avoidance), enabling analysis of communication protocol effectiveness in more diverse MARL settings.
% \end{itemize}

\textbf{C1 - High-Throughput MARL enviornment}
The system should provide a configurable, vectorised Rust environment with Python bindings, achieving at least 50K environment steps per second and supporting real-time visualisations.

\textbf{C2 - MARL Framework}
A functioning multi-agent reinforcement learning implementation with a shared actor and centralised critic capable of learning the complex environment.

\textbf{C3 - Interchangeable Communication Protocols}
At least two protocols (no communication and discrete communication) and the implementation of the AI Mother Tongue architecture, suitable for this setup, which are all interchangeable within the same training pipeline.

\textbf{C4 - Evaluation of Emergent Communication}
% A quantitative and qualitative evaluation of the communication protocols, including learned performance, codebook utilisation and semantic/pragmatic analysis.
A quantitative evaluation of the communication protocols, including achieving a Gini coefficient of at least 0.7, proving a low-entropy active vocabulary and demonstrating a Zipfian distribution in codebook utilisation.
Furthermore, the communication must exhibit context-aware signalling, verified with conditional probability distributions (e.g., $P(\text{Target}|\text{Code})$).


\textbf{C5 - Software Engineering and Reproducibility}
The framework should be reusable for other researchers.
This includes high testing coverage, logging, and checkpointing.

These criteria directly follow from the central research question from section \ref{sec:1}.
C1 and C2 ensure the question is asked in a non-toy setting.
Without the high-throughput MARL environment and the stable MARL training, the AI Mother Tongue architecture \cite{AIMotherTongue} would not be evaluated under realistic conditions.
C1 and C2 also establish a robust, highly scalable foundation for future MARL research.
C3 ensures a comparative evaluation of the AI Mother Tongue architecture against realistic baselines.
C4 allows for a fair, in-depth evaluation of the protocol.
Finally, C5 encourages testing, configuration, logging and checkpointing to create a reusable toolkit for other researchers to further this work.

\subsection{MoSCoW Analysis}
\label{sec:2.5.3}

\begin{table}[ht]
    \centering
    \caption{MoSCoW Analysis of project requirements and associated risks.}
    \label{tab:moscow_analysis}
    \small
    \begin{tabularx}{\textwidth}{@{} l X X @{}}
        \toprule
        \textbf{Priority} & \textbf{Feature} \\ \midrule
        \textbf{Must Have} & Vectorised Rust environment with PyO3 bindings \\ \addlinespace
        \textbf{} & Functioning MAPPO-based multi-agent reinforcement learning system \\ \addlinespace
        \textbf{} & At least two communication protocols, including the AI Mother Tongue architecture \\ \midrule
        \textbf{Should Have} & Real time visualisation of simulations running \\ \addlinespace
        \textbf{} & Pipeline to train autoencoders \\ \midrule
        \textbf{Could Have} & Advanced VAE varients: SQ-VAE, HQ-VAE \\ \addlinespace
        \textbf{} & A separate environment task to test agents on \\ \addlinespace
        \textbf{} & Hyperparameter tuning for all ML training loops \\ \midrule
        \textbf{Won't Have} & Non-Stationary Environment \\ \addlinespace
        \textbf{} & Modelling of network latency and packet loss \\ \bottomrule
    \end{tabularx}
\end{table}


\section{System Design \& Tool Selection}
\label{sec:2.6}

\subsection{Software Development Model}
\label{sec:2.6.1}

The project's development used iterative sprints focused on incremental value, following the Agile methodology \cite{agile_guide}.
The requirements for emergent communication in MARL were inherently exploratory.
This is due to algorithms evolving in response to evaluations such as codebook utilisation and training success.
The Agile methodology proceeded with the following phases:
\begin{center}
\textsc{Requirements} \(\rightarrow\) \textsc{Planning} \(\rightarrow\) \textsc{Implementation} \(\rightarrow\) \textsc{Testing/Evaluation} \(\rightarrow\) \textsc{Repeat}
\end{center}

\subsection{Tool Selection \& Justification}
\label{sec:2.6.2}

% After reviewing similar projects, I settled on separating my environment and my code from my controller.
Python was chosen as the programming language for the controller primarily because of its strong machine learning and data analysis libraries.
(including but not limited to NumPy \cite{numpy}, Pandas \cite{pandas}, Matplotlib \cite{matplotlib}, and Jupyter Notebooks \cite{jupyter}).
PyTorch \cite{pytorch} was chosen to develop the models due to its high performance and imperative execution, which facilitates rapid iterative prototyping.

However, Python is not a performance-oriented language and, as of this development, has a GIL \cite{python_gil_wiki}, meaning that simulations of the environment must be executed serially, severely bottlenecking performance.
Rust was developed for its high performance and memory safety, enabling rapid development.
To enable vectorisation of the environment, rayon \cite{rayon_rs} was chosen, as it provides safe, data-parallel multi-threading, enabling the simulation to concurrently step through all current worlds.

To build the real-time, lightweight visualisation frontend decoupled from the high-performance simulation, TypeScript \cite{typescript} and p5.js \cite{p5js} were chosen.
TypeScript \cite{typescript} is chosen for its strict typing to enforce the data types communicated through the websocket, ensuring consistency between the frontend and the simulation.
P5.js \cite{p5js} is used to render the visualisation, chosen for its simplicity.
For the bridge between the decoupled systems, tokio \cite{tokio} was chosen to spawn an asynchronous background thread, allowing the main environment to execute completely independently.
The warp \cite{warp} web server framework was selected to manage the WebSocket connection, ensuring reliable, low-latency streaming of state data from the simulation to the frontend.

To streamline package management and controller environment integration, I used a pyproject.toml file.
This file handled installing packages from external sources and building the Rust environment program to allow the Python controller to access it.

For version control, GitHub \cite{github} was used to store the git repository, which included both the program and the dissertation.
Regular updates were made to the code on my laptop as an additional backup.

\section{Starting Point}
\label{sec:2.7}

Prior to this project, I had extensive experience in Python and a foundational understanding of neural networks from Cambridge courses and self-instructed exercises, such as creating a neural network from scratch to solve the XOR problem and using genetic algorithms to play the game Snake.
My recent industry experience introduced me to the Rust programming language and to the Py03 library for working with both Rust and Python.
However, I had no prior experience with the PyTorch ecosystem, Variational Autoencoders, or deep Reinforcement Learning paradigms (such as PPO or MAPPO).
Consequently, a significant initial phase of this project was dedicated to learning these.
